
\documentclass[master=elt, cleveref, autoref, masteroption=eg]{kulemt}

\setup{title={Microbial communities through the lens of 
               high throughput sequencing,
               data integration and 
               metabolic networks analysis},
  author={Haris Zafeiropoulos},
  promotor={Prof. Emmanouil Ladoukakis \\ Dr Evangelos Pafilis \\ Dr Christoforos Nikolaou},
  assessor={},
  assistant={}}

% Choose the fonts for the regular text, e.g. Latin Modern
\setup{font=utopia}

% Here you can load other packages or provide your own definitions
% ----------------------------------------------------------------
% Finally, hyperref is used for PDF files.
\usepackage[pdfusetitle, colorlinks, citecolor=cyan, plainpages=false]{hyperref}
\usepackage[english,greek]{babel}
\usepackage[utf8x]{inputenc}

% To enable the \citeauthor function // there is a chance that this is why we cannot have author names 
% \usepackage[square, numbers]{natbib}
\usepackage{natbib}
\bibliographystyle{abbrvnat}

% To have multirows in tables
\usepackage{multirow}

% To write MMCS algorithm
\usepackage{algorithm, algorithmic}

% To have rotated tables
\usepackage{rotating} 

% To have check boxes
\usepackage{pifont}
\usepackage{amsmath}

% The xspace package provides a single command that looks at what comes after 
% it in the command stream, and decides whether to insert a space to replace 
% one “eaten” by the TEX command decoder. 
% The decision is based on what came after any space, 
% not on whether there was a space (which is unknowable): 
% so if the next thing proves to be punctuation, the chances are there was 
% no space, but if it’s a letter, there’s probably a need for space. 
% This technique is not perfect, but works in a large proportion of cases.
\usepackage{xspace}

% For lemas
\usepackage{amsthm}
\usepackage{amssymb}

% To have todo notes
\usepackage{todonotes}

% To have part A, B etc in a figure
\usepackage{subcaption}
\usepackage{caption}

% to have multiline equations
\usepackage{mathtools}

% DEFINITIONS
\def\RR{{\mathbb R}}

% COMMANDS
\newcommand{\dd}[1]{\mathrm{d}#1}
\newcommand{\OO}{\mathcal{O}\xspace}
\newcommand{\sO}{\widetilde{\mathcal{O}}\xspace}
\newcommand{\myparagraph}[1]{\paragraph{#1}\mbox{}\\}

\newtheorem{definition}{Definition}

% Set the depth of Table of Contents
\setcounter{tocdepth}{3}


%Begin the document 
\begin{document} 

% PREFACE
\begin{preface}


  
   % \textit{
   %    "The problem is to construct a third view, one that sees the entire world 
   %    neither as an indissoluble whole nor with the equally incorrect, but currently dominant, 
   %    view that at every level the world is made up of bits and pieces that can be isolated 
   %    and that have properties that can be studied in isolation. 
   %    Both ideologies [...] prevent a rich understanding of nature and prevent us from solving 
   %    the problems to which science is supposed to apply itself."
   % }\\
   % \hspace*{\fill} \textbf{Lewontin}, Biology as Ideology \\

   % \textit{
   %    "
   %    Thus at every step we are reminded that we by no means rule over nature like a conqueror over a foreign people, 
   %    like someone standing outside nature - but that we, with flesh, blood and brain, belong to nature, 
   %    and exist in its midst, and that all our mastery of it consists in the fact that we have the advantage over 
   %    all other creatures of being able to learn its laws and apply them correctly." \\
   % } \\
   % \hspace*{\fill}  \textbf{Friedrich Engels}, Dialectics of Nature 

   % \noindent\hfil\rule{0.5\textwidth}{.4pt}\hfil \\
 
   This thesis is the product of a series of chance-and-necessity events.
   My home village, my father and a book
   \footnote{\href{https://www.simonandschuster.com/books/My-Life-as-an-Indian/J-W-Schultz/9781634504270}{\textit{My life as an Indian}}, by J. W. Schultz (1907)} 
   made me appreciate nature and think highly of the relationship between humans and their environment. 
   The first two, they also made me, among other things, rather stubborn. 
   This stubbornness led me to first join the Technical University of Patras,
   even if Biology was my passion since secondary school,
   but then, somehow back to biology and the University of Athens. 
   Yet, it was only chance that got me in the city that shaped me the most.
   A few months before defending my bachelor's thesis,
   \href{http://ebd10.ebd.csic.es/}{Prof. Pedro Jordano},
   who has contributed the most to the study of ecological networks and interactions,
   pointed out to me 
   that I need first to get some expertise in Bioinformatics if I am interested in his work.
   This was the first time I ever considered of Bioinformatics.
   It was the same year that a relative MSc in the University of Crete ran for the first time. 
   % After exhausting everyone's patience
   % the program aggreed to allow me to give an online interview
   % and after a student said she would not join the MSc, I moved in Heraklion. 
   A few years later and mostly thanks to \href{https://cpavloud.github.io/mysite/}{Dr. Christina Pavloudi}  
   it is hard to tell \textit{what if} I was not studying microbial assemblages.

   Therefore, it is all the good and bad circumstances and choices as well as the people 
   related to each of those 
   that brought me here today, 
   and it is them
   that I would like to acknowledge first.

   I would like to thank my promotors;
   \href{https://ladoukakis.weebly.com/}{Prof. Manolis Ladoukakis}, 
   a university teacher 
   with whom we met
   when I invaded his office crying for help and he just said
   \textit{"breathe; we will figure this out"}. 
   Since then, I enjoy his guidance for which I am grateful. 
   \href{http://lab42open.hcmr.gr/}{Dr. Evangelos Pafilis} with whom we started working together back in 2017 in 
   the framework of my MSc thesis 
   and 6 years later, here we are. 
   % I would not have joined this PhD program if it was not for him
   % and most probably I would have not stayed in academia in general.
   % It was him who made me realise the value of metadata and the importance of the community-driven 
   % efforts to address several challenges on bioinformatics. 
   \href{http://computational-genomics.weebly.com/}{Dr. Christoforos Nikolaou} was also among our MSc teachers
   and has been an influence to me since I first came in Crete.
   % moreover, being a Liverpool fan, a bibliophagist and a writer himslef 
   % has played a great part in making me remember to always keep the balance. 
   Also, \href{https://cpavloud.github.io/mysite/}{Dr. Christina Pavloudi} with whom we met during my MSc and since then, she has
   not stopped triggering my curiosity and has (almost) never complained with my endless questions to her.
   Special thanks to~\href{https://tolischal.github.io/}{Dr. Apostolos Chalkis}, 
   probably the greatest example of a necessity-and-chance case in my research story so-far, 
   with whom we met on the streets of Athens 
   and a few years later he called me to ask me if I had ever heard of metabolic networks; 
   this was the first time I ever heard of them.
   \href{https://who.paris.inria.fr/Elias.Tsigaridas/}{Prof. Elias Tsigaridas} 
   and~\href{https://who.paris.inria.fr/Elias.Tsigaridas/}{Dr. Vissarion Fisikopoulos} 
   thank you for your patience and spirit; 
   I learnt more than a lot working with the \href{https://geomscale.github.io/}{GeomScale} group. 
   % and it is my belief we were all 
   % astonished with how magnificent science can be when merging fields and exploring fields you have no idea about. 
   I would also like to say a great thank you to the~\href{https://sites.google.com/view/lab-area52/home}{Area52} lab. 
   To \href{https://sites.google.com/view/lab-area52/home/meet-the-team?authuser=0#h.p_TlJiVLurzMia}{Prof. Jens Carlsson} 
   who convinvced me to look for "aliens" and proved me there are people that have no idea
   about how a good olive oil tastes like! 
   To \href{https://www.researchgate.net/profile/Sanni-Hintikka}{Dr. Sanni Hintikka}
   and to Dr. Laura Gargan that would always make some time for me, even from the back seat of a car during lunch time.
   Last but not least, \href{https://imbbc.hcmr.gr/user/kotoulas/}{Dr. George Kotoulas} who has been an inspiration to me from so many and different aspects.

   I would also like thank and fairwell all members of the Biodiversity lab and colleagues in HCMR. 
   \foreignlanguage{greek}{Χλαπάτσα}, Niki, Eva, Emma, Eirini, George, Bill, Antoni, Stelio, Natassa, Noche Sangre, Adrianouko, Eirini, Despoina, 
   thank you for making the every-day-life so companionable; I will miss you all.
   The most special thanks to 
   \href{https://www.researchgate.net/profile/Savvas-Paragkamian-2}{Savvas Paragkamian};
   over the last years, we shared a desk, a great number of issues and bugs, our ideas and our temper.
   % I would like to believe that more than a great colleague, I also gained a friend.
   % kept reminding me why I am doing science even in my darkest hours. 

   % Eva, Niki (\foreignlanguage{greek}{τσιγαράκι};), George, Bill,  
   % Speaking of voices, Wanda thanks for being soo Wanda to all of us! 
   % You are our rock!
   % Emma and Eirini I would consider becoming a web developer only to work with you. 
   % And of course, 
   % Especially to our admins Stelio and Antoni, the best sys-admins ever as well as among the greatest aficionados of \foreignlanguage{greek}{γουρουνοπούλα}.
   % \foreignlanguage{greek}{Χλαπάτσα} you goo, I tried so hard not to mention you but you found your way in;
   % thank you for being \textit{the only voice of reason}.
   % Also, Georgia, Eirini and Despoina that have made the PhD life so much bearable and fun. 


   % Along with our tutors 
   % \href{https://www.ics.forth.gr/cbml/person/kanterakis/alexandros%C2%A0}{Dr. Alexandros Kanterakis}, 
   % \href{https://scholar.google.com/citations?user=HNOIZrAAAAAJ&hl=en}{Dr. Spiros Chavlis}, 
   % \href{http://pop-gen.eu/wordpress/}{Pavlos Pavlidis}, 
   % \href{https://economics.soc.uoc.gr/en/staff/37351/17}{Prof. Michalis Tsagris} 
   % % (typical example of a nice guy from the \textit{"holly grounds"} of Greece) 
   % you have been such a driving force to me.
   % all of them, each one on his/her own way, 
   % pointed out aspects of our science I would have never think on my own. 
   % I have to admit that there is always a sly smile on my face every time I think of them 
   % and I consider myself privileged having most of them in my corner. 
   % Pascha and Mitso for all those "\foreignlanguage{greek}{ε, ναι ρε φίλε!}". 
   % To Kleioudi for learning me what a \textit{string} variable is and so much more since then.

   % Since high school and the December of 08', it has been a rather 
   % Sars-Cov-2 revealed so many of the pathogenies of our societies.
   Last, I would like to declare my respect and gratitude to all those
   that set the example for me at every turn.
   At a collective level, the~\href{https://nka.gr/ti-einai-i-nka/}{Communist Liberation youth}; 
   it might gets tough from time to time, but struggling to 
   interpret this world and fighting to change it,
   is most nobel fight one might fight. 
   And of course, to my corner;
   Tsocha, Nef,
   % for helping me filter what counts and what does not. 
   % Nouso, %and Mougia that some years ago \textit{"scientifically"} demonstrated that 
   % a mathematician, a physicit and a biologist can be more than characters in a joke.
   % Also, for never giving up on fighting for what is right. 
   Angelique,
   Annoula, %for being my compass, 
   % pointing out what's right and what's wrong; most of the times without knowing it.
   % Vali, %for always bringing us all closer to reality; except for the times you go nuts.
   Leo,
   \foreignlanguage{greek}{καλύτερα της πιάτσας},
   in the most various ways, you have been my \textit{lee side}. 
   I love you all and you are in my heart.
   \foreignlanguage{greek}{Σκουπίδια} this applies for all of you too! 
   % We will accomplished the merging of our scientific fields some day I am sure!
   My parents that have tought me that there is nothing common in~\textit{common people}.
   I am proud of you and there is no way I can thank you enough. 
   My sister, my super hero. 
   Last, \foreignlanguage{greek}{\textit{στο δίπλα μαξιλάρι αριστερά/ κοιμάται ο άνθρωπος ο φίλος και η λύση μου}}. 
   \foreignlanguage{greek}{Αφούλη} thank you for being my friend, my inspiration, 
   my boxing bag, my colleague, my courage, my caress, my coffee and my tobacco. \\

   \foreignlanguage{greek}{
      \hspace*{\fill} Που πας; άκουσες ποτέ την κραυγή της θάλασσας;  \\
      \hspace*{\fill} ... \\
      \hspace*{\fill} Και μέσα στα ανοιχτά με σκέπασε ένα κίνημα\\
      \hspace*{\fill} σκόρπια διαδήλωση και δύσκολο ξεκίνημα\\
      \hspace*{\fill} και σου ζητάω συγγνώμη αν λίγο τη χάλασα\\
      \hspace*{\fill} μα η διαδήλωση μια μέρα έγινε θάλασσα\\
      % \hspace*{\fill} ας πούμε πως λίγο πιο δυνατά τ’ αγκάλιασα\\
      % \hspace*{\fill} μα αυτή τη θάλασσα με τίποτα δεν άλλαζα\\ 
      % \hspace*{\fill} γιατί μου δώσανε μια πόλη και την άλωσα\\
      % \hspace*{\fill} ένα κελί που μπήκα μέσα και μεγάλωσα\\
   }

   \noindent\hfil\rule{0.5\textwidth}{.4pt}\hfil \\
 
   To a little seed, meant to join us any time now

   \foreignlanguage{greek}{Στον νεότερο από τους σποράκους που όπου να'ναι φτάνει}


\end{preface}

% TOC
\tableofcontents
\newpage

%  Abstract in english 
\begin{abstract}

   % THE QUESTION 
   % ------------
   Microbial communities are a cornerstone for most ecosystem types.
   % From marine sediments to the human gut, microbial communities impact both the structure and the 
   % functions of an ecosystem, through their abundance, versatility and interactions. 
   To elucidate the mechanisms governing such assemblages, it is fundamental to identify 
   the taxa present (\textit{who}) 
   and the processes that occur (\textit{what}) 
   in the various environments (\textit{where}). 
   Thanks to a series of technological breakthroughs vast amounts of information/data from all the 
   levels of the biological organization
   % i.e. genome, transcriptome, proteome, metabolome 
   have accumulated over the last decades.
   In this context, microbial ecology studies are now relying on bioinformatics methods and analyses.
   Therefore, a great number of challenges both from the biologist- and the computer scientist point-of-view 
   have arisen; 
   one among the most emerging ones being: 
   \textit{"what shall we do with all these pieces of information?"}. 
   The paradigm of Systems Biology addresses this challenge by moving from reductionism
   to more holistic approaches attempting to interpret how the properties of a system emerge. 

   Aim of this PhD was to 
   enhance microbiome data analyses by developing software addressing on-going computational challenges 
   on the study of microbial communities.
   On top of that, to exploit such state-of-the-art methods to study microbial assemblages in 
   extreme environments.
   To this end, the Tristomo marsh in Karpathos island (Greece), was chosen as a study case.

   % THE APPROACH 
   % -------------
   % HTS
   Environmental DNA and metabarcoding have been widely used to estimate the biodiversity (the \textit{who})
   and the structure of communities. 
   Vast amount of sequencing data targeting certain marker genes depending on the taxonomic group of 
   interest became available thanks to High Throughput Sequencing technologies. 
   However, the bioinformatics analysis of such data require multiple steps and parameter settings as well as increased computing resources. 
   Workflows along with computing infrastructures ease this need to a great extent;
   in this notion, a Pipeline for environmental DNA Metabarcoding Analysis (\texttt{PEMA}) was developed (Chapter~\ref{publ:pema}). 
   However, eDNA metabarcoding has limitations too. 
   Cytochrome c oxidase subunit I (COI) marker gene is a commonly used marker gene, especially in studies targeting eukaryotic taxa.
   It is well known that in COI studies a great number of the derived Operational Taxonomic Unitss (OTUs) get no taxonomic hits. 
   The presence of pseudogenes but also of non-eukaryotic taxa among the amplicon data, 
   with the simultaneous absence of the latter from the most commonly-used reference databases 
   justify this phenomenon to a great extent. 
   To identify such cases the Dark mAtteR iNvestigator~(\texttt{DARN}) software was developed; 
   \texttt{DARN} makes use of a COI-oriented tree of life to provide further insight to such
   known unknown sequences (Chapter~\ref{publ:darn}).

   % DATA INTEGRATION
   Amplicon and shotgun metagenomics approaches 
   % satisfy the question of \textit{who} is present in a community.
   % The latter, as well as other omics technologies, 
   % provide great insight into the question of \textit{what} processes occur in a smaple. 
   along with the rest of the omics technologies, 
   have led to vast amount of data and metadata, 
   recording the \textit{who}, the \textit{what} and the \textit{where}. 
   To enable optimal accessibility and usage of this information,
   a great number of databases, ontologies as well as community-standards 
   have been developed.
   By exploiting data integration techniques to bring such bits of information together, 
   as well as text mining methods to extract knowledge "hidden" among the billions of text lines in 
   already published literature, the PREGO knowledge-base returns thousands of 
   \textit{what - where - who} potential associations (Chapter~\ref{cha:prego}).

   % MET NETS
   The driving question though is \textit{how} the different microbial taxa ascertain their 
   endurance as part of a community. 
   Metabolic interactions among the various taxa play a decisive role for 
   the composition of such assemblages. 
   Genome-scale metabolic networks (GEMs) enable the inference of such interactions.
   Random sampling on the flux space of such metabolic models, 
   provides a representation of the flux values a model can get under various conditions.
   However, flux sampling is challenging from a computational point of view, especially 
   as the dimension of a metabolic model increases.
   % Markov Chain - oriented approaches can provide alternatives to overvome these challenges. 
   To address such challenges, a Python library called \texttt{dingo} was developed 
   using a Multiphase Monte Carlo Sampling algorithm (Chapter~\ref{chap:dingo}).
   % GEMs were built using the MAGs retrieved from the Tristomo swamp and 
   % metabolic interactions between them and their environment were investigated.   

   % AFOULO-SWAMP
   Finally, sediment and microbial mat samples as well as microbial aggregates from 
   a hypersaline marsh in Tristomo bay (Karpathos, Greece) were analyzed. 
   Both amplicon (16S rRNA) and shotgun sequencing data were used to characterize 
   the microbial structure of the communities
   and environmental parameters (e.g. salinity, oxygen concentration) 
   were measured at the sampling sites.
   Key functions supporting life in such environments were identified and metagenome-assembled genomes (MAGs)
   of novel species found were built
   (Chapter~\ref{publ:swamp}). 


   % A SHORT CONLCUSION 
   % ------------------
   Similar to microbial communities, bioinformatics methods tend to build assemblages while 
   "living" on your own is quite rare. 
   The methods developed during this PhD project combined with state-of-the-art methods
   anticipate to build a framework  that enables
   moving from the community to the species level and then back again to the one of the community. 
   Such a framework is described for the study of microbial interactions at real-world communities. 

\end{abstract}

%  Abstract in greek
\begin{abstract*}


   Οι μικροβιακές κοινότητες αποτελούν ακρογωνιαίο λίθο για τους περισσότερους τύπους οικοσυστημάτων.
   Για να διευκρινιστούν οι μηχανισμοί που καθορίζουν τέτοιες κοινότητες είναι καθοριστικής σημασίας 
   η αναγνώριση των τάξων που τις απαρτίζουν (\textit{ποιος}) καθώς και των διεργασιών που πραγματοποιούνται (\textit{τι})
   στους διάφορους τύπους περιβαλλόντων (\textit{που}).
   Χάρη σε μια σειρά τεχνολογικών επιτευγμάτων, ιδιαίτερα μεγάλες ποσότητες πληροφορίας/δεδομένων 
   από όλα τα επίπεδα οργάνωσης της ζωής έχουν σωρευτεί τις τελευταίες δεκαετίες. 
   Σε αυτό το πλαίσιο, οι μελέτες μικροβιακής οικολογίας είναι άρρηκτα συνδεδεμένες και βασίζονται σε βιοπληροφορικές μεθόδους και αναλύσεις. 
   Ωστόσο, έχει προκύψει ένας σημαντικός αριθμός προκλήσεων τόσο από την βιολογική σκοπιά όσο και από αυτήν την επιστήμης υπολογιστών.
   Μεταξύ αυτών, καθοριστικό ερώτημα αποτελεί το \textit{τι μπορούμε να κάνουμε με όλα αυτά τα επιμέρους κομμάτια πληροφορίας;}.
   Το παράδειγμα της Βιολογίας Συστημάτων απαντά σε αυτό το ερώτημα περνώντας από πιο αναγωγικές σε πιο ολιστικές προσεγγίσεις
   προσπαθώντας να ερμηνεύσει το πως προκύπτουν και συνδέονται οι ιδιότητες ενός συστήματος. 

   Στόχος αυτής της διδακτορικής διατριβής ήταν να ενισχύσει την ανάλυση δεδομένων από μικροβιώματα 
   αναπτύσσοντας λογισμικά εργαλεία που να απαντούν σε τρέχουσσες υπολογιστικές προκλήσεις για την 
   μελέτη μικροβιακών κοινοτήτων. 
   Επιπλέον, να μελετήσει μικροβιακές κοινότητες σε ακραία περιβάλλοντα εφαρμόζοντας σύγχρονες μεθόδους 
   για την αναγνώριση τάξων και διεργασιών. 
   Για την επίτευξη αυτού του στόχου, το έλος Τριστόμου στο νησί της Καρπάθου, επιλέχθηκε ως περιοχή μελέτης.

   Το περιβαλλοντικό  \foreignlanguage{english}{DNA}  και η μέθοδος της μετακωδικοποίησης έχουν χρησιμοποιηθεί σημαντικά για την εκτίμηση της βιοποικιλότητας
   (\textit{ποιος}) και τη δομή των κοινοτήτων. 
   Σημαντικός αριθμός αλληλουχικών δεδομένων που στοχεύουν σε ορισμενα γονίδια δείκτες και που εξαρτόνται από τις ταξινομικές ομάδες στόχους, 
   είναι διαθέσιμα χάρη στις τεχνικές αλληλούχισης υψηλής απόδοσης \foreignlanguage{english}{HTS}. 
   Ωστόσο, η βιοπληροφορική ανάλυση τέτοιων δεδομένων απαιτούν μεγάλο αριθμό βημάτων και παραμέτρων καθώς και σημαντικούς υπολογιστικούς πόρους. 
   Οι ροές εργασιών σε συνδυασμό με υπολογιστικές υποδομές μπορούν να απαντήσουν σε αυτές τις απαιτήσεις σε σημαντικό βαθμό. 
   Σε αυτό το πλαίσιο αναπτύχθηκε η ροή εργασίας \foreignlanguage{english}{PEMA} με στόχο την ανάλυση δεδομένων μετακωδικοποίησης από περιβαλλοντικό~\foreignlanguage{english}{DNA}.
   Κεφάλαιο~\foreignlanguage{english}{\ref{publ:pema}}.
   Ωστόσο, η μέθοδος μετακωδικοποίησης χαρακτηρίζεται από σειρά περιορισμών. 
   Η υπομονάδα~\foreignlanguage{english}{I} της κυτοχρωμικής οξειδάσης~\foreignlanguage{english}{c (COI)}, αποτελεί έναν δείκτη 
   που χρησιμοποιείται ευρέως, ειδικά στην περίπτωση ευκαρυωτικών τάξων - στόχων. 
   Είναι γνωστό πως σε μελέτες όπου ο δείκτης αυτός χρησιμοποιείται, ένας μεγάλος αριθμός των λειτουργικών 
   ταξινομικών μονάδων \foreignlanguage{english}{(OTUs)} που προκύπτουν, δεν καταφέρνουν να ταυτοποιηθούν. 
   Η παρουσία τοσο ψευδογονιδίων όσο όμως και μη-ευκαρυωτικών τάξων ανάμεσα σε τέτοια αλληλουχικά δεδομένα,
   με την ταυτόχρονη απουσία των τελευταίων από τις βάσεις αναφοράς, εξηγεί την μη ταυτοποίησή τους σε σημαντικό βαθμό. 
   Για την αναγνώριση τέτοιων περιπτώσεων, αναπτύχθηκε το υπολογιστικό εργαλείο \foreignlanguage{english}{DARN} το οποίο 
   αξιοποιεί ένα φυλογενετικό δέντρο που καλύπτει και τις τρεις επικράτειες του δέντρου της ζωής, βασισμένο σε αλληλουχίες του 
   δείκτη~\foreignlanguage{english}{COI}, Κεφάλαιο~\foreignlanguage{english}{\ref{publ:darn}}.

   Μέθοδοι γονιδίων δεικτών και μεταγονιδιωματικής καθώς και το σύνολο των μεθόδων αλληλούχισης υψηλή απόδοσης, 
   έχουν οδηγήσει στην σώρευση σημαντικά μεγάλου αριθμού δεδομένων και μεταδεδομένων 
   καταγράφοντας τάξα και διεργασίες σε σειρά τύπους περιβαλλόντων.
   Για να επιτρέψουν την βέλτιστη προσβασιμότητα και αξιοποίηση αυτής της πληροφορίας,
   έχουν δημιουργηθεί
   σειρές βάσεων δεδομένων, οντολογιών αλλά και προτύπων - κανόνων για να ακολουθεί η κοινότητα 
   για την καταχώρηση τους.
   Αξιοποιώντας μεθόδους ενσωμάτωσης/ολοκλήρωσης δεδομένων~(\foreignlanguage{english}{data integration}) για 
   την εύρεση των διάφορων κομματιών πληροφορίας και την συσχέτισή τους, καθώς και τεχνικών εξόρυξης κειμένου
   (\foreignlanguage{english}{text mining}) για την εξαγωγή γνώσης από το σύνολο της δημόσια διαθέσιμης βιβλιογραφίας
   αναπτύχθηκε η βάση-γνώσης \foreignlanguage{english}{PREGO}, Κεφάλαιο~\foreignlanguage{english}{\ref{cha:prego}},
   η οποία επιστρέφει χιλιάδες σχέσεις μεταξύ τάξων, περιβαλλόντων και διεργασιών. 


   Καθοριστικό ερώτημα ωστόσο σε οτι αφορά τις μικροβιακές κοινότητες, αποτελεί το "πώς" τα διάφορα μικροβιακά τάξα 
   εξασφαλίζουν την θέση τους ως μέλη της κοινότητας. 
   Μεταβολικές αλληλεπιδράσεις μεταξύ των διάφορων τάξων παίζουν καθοριστικό ρόλο για την συγκρότηση 
   τέτοιων κοινοτήτων. 
   Μεταβολικά δίκτυα στην κλίμακα του γονιδιώματος (\foreignlanguage{english}{GEMs}) επιτρέπουν την 
   αναγνώριση τέτοιων αλληλεπιδράσεων. 
   Η τυχαία δειγματοληψία στον χώρο που ορίζεται από τις πιθανές τιμές που μπορεί να πάρουν οι ροές των αντιδράσεων (\foreignlanguage{english}{flux sampling})
   επιτρέπει την αναπαράσταση των τιμών που μπορεί να λάβουν αυτές οι ροές κάτω από συγκεκριμένες συνθήκες. 
   Ωστόσο η μέθοδος \foreignlanguage{english}{flux sampling} είναι ιδιαίτερα απαιτητική από υπολογιστική σκοπιά,
   ιδιαίτερα όσο η διάσταση του μεταβολικού μοντέλου αυξάνει.
   Για τον σκοπό αυτό αναπτύχθηκε η βιβλιοθήκη \foreignlanguage{english}{dingo} η οποία κάνει χρήση ενός 
   πολυφασικού αλγορίθμου \foreignlanguage{english}{Monte Carlo}, Κεφάλαιο~\foreignlanguage{english}{\ref{chap:dingo}}. 

   Τέλος, αναλύθηκαν δείγματα ιζήματος από το έλος Τριστόμου Καρπάθου, καθώς επίσης
   δείγματα από μικροβιακούς τάπητες (\foreignlanguage{english}{mat}) και από μικροβιακά συσσωματώματα~(\foreignlanguage{english}{aggregates}).
   Για τον σκοπό αυτό, χρησιμοποιήθηκε τόσο η μέθοδος μετακωδικοποίησης με γονίδιο-δείκτη το 16\foreignlanguage{english}{S}
   όσο και η μέθοδος μεταγονιδιωματικής \foreignlanguage{english}{shotgun}.
   Επίσης μετρήθηκαν περιβαλλοντικές παράμετροι (όπως αλατότητα, συγκέντρωση οξυγόνου). 
   Βασικές λειτουργίες που υποστηρίζουν τη ζωή σε τέτοιες συνθήκες εντοπίστηκαν 
   ενώ ακόμη γονιδιώματα τάξων που εντοπίζονται για πρώτη φορά ανασκευάστηκαν από τις αλληλουχίες του μεταγονιδιώματος~(\foreignlanguage{english}{MAGs}), Κεφάλαιο~\foreignlanguage{english}{\ref{publ:swamp}}.

   Όπως συμβαίνει και στις μικροβιακές κοινότητες, οι βιοπληροφορικές μέθοδοι σπάνια στέκουν 
   απομονωμένες, αντίθετα τείνουν να συγκροτούν κι αυτές τις δικές τους "κοινότητες". 
   Οι μέθοδοι που αναπτύχθηκαν στα πλαίσια αυτής της διατριβής επιδιώκουν να συγκροτήσουν ένα πλαίσιο μελέτης
   από το επίπεδο της κοινότητας σε αυτό του είδους και από εκεί, πίσω πάλι στην κοινότητα. 
   Ένα τέτοιο πλαίσιο αναλύεται για την μελέτη μικροβιακών αλληλεπιδράσεων. 



\end{abstract*}

% A list of figures and tables is optional
\listoffiguresandtables

% A section for abbreviations and symbols
\chapter{List of Abbreviations and Symbols}

% The abbrevieations mentioned
\section*{Abbreviations}
   \begin{flushleft}
   \renewcommand{\arraystretch}{1.1}
   \begin{tabularx}{\textwidth}{@{}p{12mm}X@{}}
      COI   & Cytochrome c oxidase subunit I\\
      ITS   & Internal Transcribed Spacer\\
      NGS   & Next Generation Sequencing\\
      eDNA  & environmental DeoxyriboNucleic Acid\\
      OTU   & Operational Taxonomic Unit\\
      ASV   & Amplicon Sequence Variant\\
      HPC   & High Performance Computing\\
      MAG   & Metagenome-assembled Genome\\
      MCMC  & Markov Chain Monte Carlo\\
      MMCS  & Multiphase Monte Carlo Sampling\\
      PREGO & PRocess Environment OrGanisms\\
      PEMA  & Pipeline for Environmental DNA Metabarcoding Analysis\\
      DARN  & Dark mAtteR iNvestigator\\
      PSRF  & Potential Scale Reduction Factor\\
      ESS   & Effective Sample Size\\
   \end{tabularx}
\end{flushleft}


% The symbols mentioned
\section*{Symbols}
   \begin{flushleft}
   \renewcommand{\arraystretch}{1.1}
   \begin{tabularx}{\textwidth}{@{}p{12mm}X@{}}
      $\RR$          & Set of real numbers\\
      $\OO$          & Algorithm complexity\\
      $\sO$          & Algorithm complexity ignoring polylogarithmic factors\\
      $\mathbb{E}$   & Expected Value operator\\
      % $\mathcal{U}$  & utility function....\\
      $\ell$         & ray\\
      $P$            & polytope\\
      $\tau$         & trajectory\\
      $\partial P$   & boundary of the $P$ polytope\\
      $v$            & flux vector\\
      $v_i$          & flux value of the reaction $i$\\
      $W$            & walk length\\
      $\rho$         & number of reflections\\
   \end{tabularx}
\end{flushleft}

% ---------------------------------------------------
% The main text!
% ---------------------------------------------------

% Now the actual text begins
\mainmatter

% Introduction 
\include{chapters/intro}

% The who : microbial diversity
\include{chapters/chap-1}

% The who, the what and where: PREGO
\include{chapters/chap-2}

% Flux sampling
\include{chapters/chap-3}

% Karpathos case
\include{chapters/chap-4}

% Computing in bio-research
\include{chapters/chap-5}

% Finally... 
\include{chapters/conclusion}

% bUFFy funding
\include{chapters/acknowledgements}

% If there are attachments
\appendixpage*          
\appendix
\include{appendices/app-A}
\include{appendices/app-B}
\include{appendices/app-C}

\backmatter 


% Add ref 
\bibliography{references}


% The one-page CV
\include{chapters/one-pagep-cv}

% End of document
\end{document}


% nice words I could use! 
% intricately = πολύπλοκα
% extrapolate = extend the application of (a method or conclusion, especially one based on statistics) to an unknown situation by assuming that
%               existing trends will continue or similar methods will be applicable. e.g the results cannot be extrapolated to other patient groups
% pertain     = που αναφέρονται, σχετικά με (παράδειγμα: challenges pertaining to data pre-processing)
% ample       = επαρκής 
% postulate = suggest or assume the existence, claim, αξιώ
% ascertain = diapistono 
% 